---
title: dplyr tutorial
author: Johan Junkka
date: 2015-06-03
css: style.css
output:
  html_document:
    toc: true
---

# Introduction

(from dplyr vignette)

When working with data you must:

* Figure out what you want to do.
* Precisely describe what you want in the form of a computer program.
* Execute the code.

The dplyr package makes each of these steps as fast and easy as possible by:

* Elucidating the most common data manipulation operations, so that your options are helpfully constrained when thinking about how to tackle a problem.
* Providing simple functions that correspond to the most common data manipulation verbs, so that you can easily translate your thoughts into code.

# Data: person & obs

We have two datasets

| -----------: | :-------------------     |
| person       | Static information       |
| obs          | Time varying information |

```{r}
library(dplyr)
library(skel15)

person <- tbl_df(person)
person
```

## Filter

Lets select all women born in 1800 or 1850 

```{r}
filter(person, KON == 2, FODDAT %/% 10000 == 1800 | FODDAT %/% 10000 == 1850)
```

## Arrange rows

Arrange by birthdate

```{r}
arrange(person, FODDAT)
```

Arrange in descending order

```{r}
arrange(person, desc(FODDAT))
```

## Selecting variables

Lets use a subset of the data

```{r}
select(person, ID, KON, FODDAT, DODDAT)
```

However, we might be interested in only the first 6 variables and FODDAT.

```{r}
select(person, ID:PARITET_G, FODDAT)
```

## Srtinging commands together

Using the obersvation data `obs` we can check the number of unique individuals. 

```{r}
data(obs)

obs <- tbl_df(obs)

distinct(select(obs, id)) 
```

or stringing them together in one operation using the "pipes" operator `%>%`

```{r}
obs %>% select(id) %>% distinct()
```

Now we can create more complex calculations

# Add a new column 

We might want to know if a person is observable until death

```{r}
person %>% 
  select(ID, DODDAT, UTDATMIN) %>% 
  mutate(UTDATMIN - DODDAT)
```

# Make event history data

1. Get all married women

```{r}
women <- obs %>% 
  select(id, sex, civst, enter) %>% 
  filter(sex == "woman", civst == "married") %>% 
  group_by(id) %>%
  arrange(enter) %>% 
  filter(row_number() == 1) %>% 
  select(id, marriage_age = enter)
```

2. Get date of dissolution

```{r}
women_diss <- obs %>% 
  select(id, sex, civst, enter) %>% 
  filter(sex == "woman", civst == "prev.married") %>% 
  group_by(id) %>%
  arrange(enter) %>% 
  filter(row_number() == 1) %>% 
  select(id, diss_age = enter)
```

3. Combine marriage age and age at dissolution

```{r}
women <- left_join(women, women_diss, by = "id")
```

Not all married women had a dissolution date as some observations ended before the marriage ended. 

4. Thus, we need to add age at last observation from the `person` file.

```{r}
women_last_obs <- person %>% 
  select(ID, FODDAT, UTDATMIN) %>% # id, birth date, last observation date
  mutate(last_age  = tillTid(UTDATMIN) - tillTid(FODDAT)) %>% 
  select(
    id = ID, 
    birth_date = FODDAT, 
    last_age
  )
```

5. Add to the `women` data

```{r}
women <- left_join(women, women_last_obs, by = "id")
```

6. Now, we can create last observation date

```{r}
women <- women %>% 
  mutate(end_age = min(c(50, diss_age, last_age), na.rm = TRUE)) %>% 
  select(id, start_age = marriage_age, end_age)
```

7. Add birth date

```{r}
max_obs_date <- max(person$FODDAT)

years_diff <- function(a, b){
  as.numeric(difftime(as.Date(as.character(a), "%Y%m%d"), as.Date(as.character(b), "%Y%m%d")))/365.25
}

women2 <- person %>%
  select(ID, FODDAT) %>% 
  left_join(women, ., by = c("id" = "ID")) %>% 
  rename(birthdate = FODDAT) %>%
  mutate(
    age_at_end = years_diff(max_obs_date, birthdate)
  ) %>% 
  ungroup() %>% 
  mutate(end_age = ifelse(end_age < age_at_end, end_age, age_at_end)) %>% 
  select(-age_at_end) %>% 
  filter(end_age > start_age)

```

8. As we are interested in fertility we will need observations for each born child within the obervation span. Start with getting all children, their birth dates and their mothers id.

```{r}
children <- tbl_df(person) %>% 
  select(ID, FODDAT, MID, KON) %>% 
  filter(MID %in% women$id) %>% 
  select(id = MID, date = FODDAT, ch_id = ID, sex = KON)
children
```

9. Add info on the mothers from `women` file

```{r}
children <- left_join(children, women2, by = "id")
children
```

```{r}
births <- arrange(children, id, date)
births
```

10. Make enter

```{r}
births <- mutate(births, enter = years_diff(date, birthdate))
```


11. Remove all births that occured after end age

```{r}
births <- filter(births, enter < end_age)
```

12. Add n gender

```{r}
births = mutate(births,
    girls = ifelse(sex == 1, 1, 0),
    boys = ifelse(sex == 2, 1, 0)
  )
```

13. Sum up and add parity

```{r}
births <- group_by(births, id) %>% 
  mutate(girls = cumsum(girls), boys = cumsum(boys), parity = girls + boys)
```

13. Add exit date and event

```{r}
births <- group_by(births, id) %>% 
  mutate(
    exit = lead(enter), 
    event = ifelse(is.na(exit), FALSE, TRUE),
    exit = ifelse(is.na(exit), end_age, exit))
```

# Lets check if 

```{r}
births <- ungroup(births) %>% 
  mutate(
    cohort = (birthdate %/% 100000)*10, 
    age_group = factor((enter %/% 5)*5)
  )

```


```{r}
births2 <- group_by(births, id) %>% 
  mutate(exit = exit - min(enter), enter = enter - min(enter)) %>% 
  ungroup() %>% 
  filter(enter < exit)
```


```{r, cache = T}
fit <- coxph(Surv(enter, exit, event) ~ sex + cohort + age_group + strata(parity), data = births2)
```

# Getting results from a regression model

Using the `broom` package

```{r}

library(broom)
library(knitr)

fit_tbl <- tidy(fit, expo = TRUE)

kable(fit_tbl)

glance(fit)
```

Combining results from regressions

```{r}
bladder1 <- bladder[bladder$enum < 5, ] 
fit1 <- coxph(Surv(stop, event) ~ rx + size + number, filter(bladder1, enum == 1))
fit2 <- coxph(Surv(stop, event) ~ rx + size + number, filter(bladder1, enum == 2))

(tbl_fit1 <- tidy(fit1, expo = TRUE))
(tbl_fit1 <- tidy(fit1, expo = TRUE))
```

Compare these models

```{r}
(tbl_fit1 <- tbl_fit1 %>% select(term, e1 = estimate, p1 = p.value))
(tbl_fit2 <- tbl_fit2 %>% select(term, e2 = estimate, p2 = p.value))

res <- left_join(tbl_fit1, tbl_fit2)

kable(res, digits = 3)
```

# More help

* Look at vignettes
* Learn to use help. 
    * What is the help file to plot a coxreg? = `?plot.coxreg`
* **Ask** questions on [Stackoverflow](http://stackoverflow.com/) & [Cross Validated](http://stats.stackexchange.com/)
* Look at other peoples code, on [GitHub](http://github.com) or replicable articles on [Demographic Research](http://demographic-research.org/volumes/replicable_articles.htm)
